Ant cuticle image classification using texture analysis: a comparative study

### 2022-06-24 paper/main.tex
add drp section

Our custom dataset of ant head images contains 2,499 images. 1072 samples of
rough textured ant cuticle textures comprise 43\% of the dataset. The remaining
1427 samples of smooth textured ant cuticle textures comprise 57\% of the
dataset. To handle the imbalance of the dataset, we apply undersampling for each
class for the training dataset. By using random stratified sampling, we
construct a training set with 800 images per class. The remaining images are
randomly split between test and validation, which turns out to roughly a
60\%/20\%/20\% train, test, and validation data split. With 272 rough samples
and 627 smooth samples left over after the stratified split, the test dataset
has roughly 136 rough samples and 313 smooth samples. Since these leftover
samples are split with code by 50\% there will be some rounding variance and
therefore the test dataset built at run-time will not always have exactly the
same number of samples.

@@

% \colorbox{green}{\parbox{\dimexpr\textwidth-2\fboxsep}{TODO: I suggest that at
%         least one deep learning based texture classification algorithm be
%         applied to the dataset for the comparative study. When this is the
%         case, one more subsection should be included here to describe
%         that/those algorithm(s).}}

@@

\section{Results}
\subsection{Environment}

\subsection{Evaluation results}

@@

\section{The analysis with visualization}

The results in the previous sections show that the fine-tuned ResNet models
outperform the VGG and randomly initialized ResNet models on the task of ant
head image classification. Additionally, the DRP models perform better than the
the general deep learning models VGG and ResNet on the task of ant head image
classification.The fine-tuned DRP single-layer model with auxiliary classifier
performed the best with an average F1 score of 0.92. We further analyze the
separation learned by both ResNet101 models in the following section.

@@

Based on the visual results of the t-SNE visualization, we can see that the
fine-tuned model learned a stronger separation of the two classes, which
reinforces the results that the fine-tuned model received a higher average
accuracy.

\subsection{GradCAM visualization}

>>>>

@@

\subsection{Deep texture analysis method used for the experiments}

In addition to the classical texture analysis methods and the deep learning
models for general image analysis, we also examine deep learning analysis
methods specifically designed for textural images. For this examination, we
adopt the deep residual pooling (DRP) network developed by Mao et al.
\cite{mao_deep_2021}. The DRP framework consists of a unique residual pooling
layer, which is formed by a residual encoding module followed by an aggregation
module. The residual encoding module extracts relevant spatial information,
while the aggregation module performs averaging to obtain orderless
low-dimension features for classification.

According to \cite{mao_deep_2021}, the DRP network can be used either with a
single residual pooling layer, applied right before the classifier layer, or
with multiple residual pooling layers, applied to the output of multiple
convolutional layers before feeding the combined pooling results into the
classifier layer. Additionally, an auxiliary classifier layer may also be used.
In our experiments, we follow these strategies. For each scenario, we conduct
two experiments. One experiment employs weights that are randomly initialized,
and the other uses weights that are fine tuned over those from a pretrained
model.

\section{Results and analysis}
\subsection{Environment}

@@

\subsection{Experimental results}

For the experiments using the deep learning algorithms, we use our custom
dataset of ant head images which contains 2,499 images. 1072 samples of rough
textured ant cuticle textures comprise 43\% of the dataset. The remaining 1427
samples of smooth textured ant cuticle textures comprise 57\% of the dataset. To
handle the imbalance of the dataset, we apply undersampling for each class for
the training dataset. By using random stratified sampling, we construct a
training set with 800 images per class. The remaining images are randomly split
between test and validation, which turns out to roughly a 60\%/20\%/20\% train,
test, and validation data split. With 272 rough samples and 627 smooth samples
left over after the stratified split, the test dataset has roughly 136 rough
samples and 313 smooth samples. Since these leftover samples are split with code
by 50\% there will be some rounding variance and therefore the test dataset
built at run-time will not always have exactly the same number of samples.

@@

\FloatBarrier
The results in the previous sections show that the fine-tuned ResNet models
outperform the VGG and randomly initialized ResNet models on the task of ant
head image classification. Additionally, the DRP models perform better than the
the general deep learning models VGG and ResNet on the task of ant head image
classification.The fine-tuned DRP single-layer model with auxiliary classifier
performed the best with an average F1 score of 0.92. We further analyze the
separation learned by both ResNet101 models in the following section.

@@

\FloatBarrier
Based on the visual results of the t-SNE visualization, we can see that the
fine-tuned model learned a stronger separation of the two classes, which
reinforces the results that the fine-tuned model received a higher average
accuracy.

\FloatBarrier
\subsection{GradCAM visualization}
